---
title: "InStep - data explanatory"
author: "Nataliya Peshekhodko"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read data


```{r, message=FALSE, warning=FALSE}
library(readxl)
library(dplyr)
library(ggplot2)
library(gbm)
library(caret)
library(randomForest)
```


Reading completed dimensions, completed data investigation tasks and demographic datasets.

```{r}
demographic_path = "data/User Demographics Filtered.xlsx"
dimensions_completed_path = "data/Dimensions Completed.xlsx"
data_investigation_completed_path = "data/Data Investigations Completed Filtered.xlsx"


user_demographic_df = read_excel(demographic_path)
dimensions_completed_df = read_excel(dimensions_completed_path)
data_investigation_completed_df = read_excel(data_investigation_completed_path)

head(user_demographic_df)
head(dimensions_completed_df)
head(data_investigation_completed_df)
```

### Analyzing public users only and completed modules


Unique combinations of 


```{r}
unique_combinations <- unique(subset(dimensions_completed_df, select = c(dimension, module)))
print(unique_combinations)
```


Looking at public users only

```{r, warning=FALSE, message=FALSE}
modules_public_users_df = dimensions_completed_df %>%
  filter(#is.na(module_completed_at) &
         !organization_name %in% c("InSTEP & RTI", "Advisors", 'First 75')) %>%
  select(user_id, organization_name, dimension, module, module_completed_at) %>%
  distinct()

head(modules_public_users_df)
```


If for the same `user_id`, `organization_name`, `dimension` and `module` there are several records and one of them has `completed_at` not `NA`, keep that value, otherwise keep `NA` and remove duplicates.

```{r}
modules_public_users_filtered_df = modules_public_users_df %>%
  group_by(user_id, organization_name, dimension, module) %>%
  mutate(module_completed_at = if_else(any(!is.na(module_completed_at)), 
                                       first(module_completed_at[!is.na(module_completed_at)]), 
                                       NA_character_)) %>%
  ungroup()

modules_public_users_filtered_df = distinct(modules_public_users_filtered_df)
modules_public_users_filtered_df
```





```{r}
user_demographic_fileterd_df = user_demographic_df %>%
  select(user_id, organization_name, country, state, hispanic, reg_grade_levels, educator_roles, teach_subjects, racial_identieis, education_level, gender)
user_demographic_fileterd_df
```

```{r}
merged_df = merge(modules_public_users_filtered_df, 
                  user_demographic_fileterd_df, by = c("user_id", "organization_name"), all.x = TRUE)
merged_df
```


Checking for missing values

```{r}
missing_values = sum(is.na(merged_df))
missing_values_per_column = colSums(is.na(merged_df))

cat("Total missing values in the dataframe:", missing_values, "\n")
print(missing_values_per_column)
```


```{r, message=FALSE}
module_counts <- merged_df %>%
  group_by(module, has_completion = !is.na(module_completed_at)) %>%
  summarise(count = n())

ggplot(module_counts, aes(x = module, y = count, fill = has_completion)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "", y = "Count", fill = "Completion") +
  ggtitle("Count of Records by Module and Completion Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1))
```



```{r, message=FALSE}

module_counts <- merged_df %>%
  group_by(education_level, has_completion = !is.na(module_completed_at),) %>%
  summarise(count = n())

ggplot(module_counts, aes(x = education_level, y = count, fill = has_completion)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Education Level", y = "Count", fill = "Completion Status") +
  ggtitle("Count of Completed and Non-completed Modules by Education Level") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

```{r}
module_counts <- merged_df %>%
  group_by(educator_roles, has_completion = !is.na(module_completed_at)) %>%
  summarise(count = n(), .groups = "drop")

ggplot(module_counts, aes(x = educator_roles, y = count, fill = has_completion)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Educator Roles", y = "Count", fill = "Completion Status") +
  ggtitle("Count of Completed and Non-completed Modules by Educator Roles") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```



```{r, message=FALSE, fig.height = 12, fig.width = 15}
module_counts <- merged_df %>%
  group_by(teach_subjects, has_completion = !is.na(module_completed_at)) %>%
  summarise(count = n(), .groups = "drop")

ggplot(module_counts, aes(x = teach_subjects, y = count, fill = has_completion)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Teaching Subjects", y = "Count", fill = "Completion Status") +
  ggtitle("Count of Completed and Non-completed Modules by Teaching Subjects") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


Create `target` variable: 0 if user didn't complete module and 1 if module was completed.

```{r}
merged_df$target = ifelse(is.na(merged_df$module_completed_at), 0, 1)
print(merged_df)
```

Checking for `NA` values again

```{r}
missing_values = sum(is.na(merged_df))
missing_values_per_column = colSums(is.na(merged_df))

cat("Total missing values in the dataframe:", missing_values, "\n")
print(missing_values_per_column)
```

```{r}
unique(merged_df$country)
```




```{r}
unique(merged_df$educator_roles)
```


```{r}
module_counts = table(merged_df$module)
print(module_counts)
```

Looks like for some modules number of observations is very small.

```{r}
educator_roles_counts <- merged_df %>%
  count(educator_roles, name = "Freq", sort = TRUE, na.rm = FALSE) %>%
  mutate(educator_roles = ifelse(is.na(educator_roles), "NA", as.character(educator_roles))) %>%
  select(-na.rm)
educator_roles_counts
```
For `educator_roles` NA values could be replaced with `Other`. 


```{r}
teach_subjects_counts <- merged_df %>%
  count(teach_subjects, name = "Freq", sort = TRUE, na.rm = FALSE) %>%
  mutate(teach_subjects = ifelse(is.na(teach_subjects), "NA", as.character(teach_subjects)))
teach_subjects_counts
```


Looks like for `teach_subjects` NA also could be replaced with `Other`. 

```{r}
education_level_counts <- merged_df %>%
  count(education_level, name = "Freq", sort = TRUE, na.rm = FALSE) %>%
  mutate(education_level = ifelse(is.na(education_level), "NA", as.character(education_level)))
education_level_counts
```

For education level `NA` values will be replaced with `Not provided`.


```{r}
gender_counts <- table(merged_df$gender, useNA = "ifany")
gender_counts
```




Drop columns `module_completed_at`, `state`, `hispanic`, `racial_identieis`, `user_id` and `reg_grade_levels`  
For column `educator_roles` replace `NA` with `Other`.
For column `teach_subjects` replace `NA` with `Other`.
For column `education_level` replace `NA` with `Not provided`. 
For column `gender` replace `NA` with `Prefer not to answer`.

```{r}
merged_df_updated <- merged_df %>%
  select(-module_completed_at, -state, -hispanic, -racial_identieis, -reg_grade_levels, -user_id, -organization_name) %>%
  mutate(
    educator_roles = ifelse(is.na(educator_roles), "Other", educator_roles),
    teach_subjects = ifelse(is.na(teach_subjects), "Other", teach_subjects),
    education_level = ifelse(is.na(education_level), "Not provided", education_level),
    gender = ifelse(is.na(gender), "Prefer not to answer", gender)
  )


print(merged_df_updated)
```

```{r}
missing_values = sum(is.na(merged_df_updated))
missing_values_per_column = colSums(is.na(merged_df_updated))

cat("Total missing values in the dataframe:", missing_values, "\n")
print(missing_values_per_column)
```


```{r}
str(merged_df_updated)
```

```{r}
merged_df_updated <- merged_df_updated %>%
  mutate_if(is.character, as.factor)

# Print the structure of the updated dataframe
str(merged_df_updated)
```


Attempt to fit XGBoost model. XGBoost is not converging.

```{r, message=FALSE, warning=FALSE}
merged_df_updated$target <- as.factor(merged_df_updated$target)

set.seed(123)
trainIndex = createDataPartition(merged_df_updated$target, p = 0.7, list = FALSE)
train = merged_df_updated[trainIndex, ]
test = merged_df_updated[-trainIndex, ]

gbm_model <- train(target ~ ., data = train, method = "gbm",
                   trControl = trainControl(method = "cv", number = 5),
                   metric = "Accuracy",
                   distribution = "bernoulli", n.trees = 100, interaction.depth = 3)

print(gbm_model)
```


```{r, message=FALSE, warning=FALSE}
rf_model = randomForest(target ~ ., data = train, ntree = 500, importance = TRUE)
print(rf_model)

test_predictions_rf = predict(rf_model, newdata = test)
test_accuracy_rf = sum(test_predictions_rf == test$target) / length(test$target)
print(paste("Test Accuracy (Random Forest):", test_accuracy_rf))
```

```{r}
print(importance(rf_model))
```
